\documentclass[en]{../../../eplsummary}

\hypertitle{cloud-INGI2145}{7}{INGI}{2145}
{Houtain Nicolas}
{Canini Marco}

\section{Introduction}

\subsection{Computing at scale}

Cloud need for scalability because modern application require
huge amounts of processing and data. Cluster (room-sized) and datacenter 
(building-sized) can provide the resources needed.
They are composed of \textbf{rack} which is a aggregation of storage devices, many
nodes and switch to connect nodes together. Unfortunately, they are not perfect.
\begin{enumerate}
    \item Difficult to dimension because they must be provisioning for the peak load
    \item Expensive in hardware invest, expertise (ex: special software) and maintenance
    \item Difficult to scale because adding new machines is not easy
\end{enumerate}

\subsection{Cloud computing}

\begin{center}
\textit{Cloud computing is a model for enabling convenient,  
on-­demand network access to a shared pool of configurable  
computing resources (e.g., networks, servers, storage,  
applications, and services) that can be rapidly provisioned  
and released with minimal management effort or  
service provider interaction.}
\end{center}

$\rightarrow$ On-demand self service, Broad network access, Resource pooling, Rapid elasticity, Measured service.

\subsubsection{Model}
Cloud computing is a business models where everything is a service:
\begin{itemize}
    \item SaaS: Software as a service
    \item PaaS: Platform as a service
    \item IaaS: Infrastructure as a service
\end{itemize}

\subsubsection{Types}
There also have three types of cloud : 
\begin{itemize}
    \item Public: commercial commercial service open to almost anyone.
    \item Community: shared by several similar organization
    \item Private: shared within a organization
\end{itemize}
In this course we focus on public cloud.

\subsubsection{Applications}

Typically, applications that involve large amounts of computation,  storage,
bandwidth Especially when lots of resources are needed quickly or load varies
rapidly.


\subsection{Virtualization}
%TODO slide 85 for 01



\section{Paper}



\end{document}
