\section{}
\subsection{}
\begin{solution}
  \begin{itemize}
    \item
      Let $\tilde{\Pi} = \langle \tilde{\Gen}, \tilde{\Mac}, \tilde{\Vrfy} \rangle$, defined as:
      \begin{itemize}
        \item $\tilde{\Gen}$: chooses a random $f$.
        \item $\tilde{\Mac}$: on input $m$, outputs $f(m)$.
        \item $\tilde{\Vrfy}$: on input $(m,t)$, outputs $1$ iff $f(m) = t$.
      \end{itemize}

      Let's analyse the maximum value of $\Pr[\MacForge_{\A, \tilde{\Pi}}(n) = 1]$ for an adversary $\mathcal{A}$.
      If after $q$ different queries (it gains no info doing the same query twice),
      $m_1, \ldots, m_q$, $\A$ outputs $(m, t)$, what are its chances of success ?
      Let $f:\{0,1\}^n \to \{0,1\}^n$.
      There are $(2^n)^{2^n}$ different $f$ and we pick a random one uniformly.
      However, there are only $(2^n)^{2^n-q}$ experiments such that $\A$ have received $(m_i,t_i)$ for $i = 1, \ldots, q$ because
      there are $(2^n)^{2^n-q}$ $f$ such that $f(m_i) = t_i$ for $i = 1, \ldots, q$.
      We could be in any of them.
      Among them, only $(2^n)^{2^n-(q+1)}$ are such that $f(m) = t$.
      Since $f$ is selected uniformly, we have

      \begin{align*}
        \Pr[\MacForge_{\A, \tilde{\Pi}}(n) = 1]
        & = \Pr[f(m) = t | f(m_i) = t_i, \forall m = 1, \ldots, q]\\
        & = \frac{\Pr[f(m) = t, f(m_i) = t_i, \forall m = 1, \ldots, q]}{\Pr[f(m_i) = t_i, \forall m = 1, \ldots, q]}\\
        & = \frac{\frac{(2^n)^{2^n-(q+1)}}{(2^n)^{2^n}}}{\frac{(2^n)^{2^n-q}}{(2^n)^{2^n}}}\\
        & = \frac{(2^n)^{2^n-(q+1)}}{(2^n)^{2^n-q}}\\
        & = \frac{1}{2^n}.
      \end{align*}
      A shortcut would have been to argued that since, $f(m)$ is independent of the $f(m_i)$ so

      \begin{align*}
        \Pr[\MacForge_{\A, \tilde{\Pi}}(n) = 1]
        & = \Pr[f(m) = t | f(m_i) = t_i, \forall m = 1, \ldots, q]\\
        & = \Pr[f(m) = t]\\
        & = \frac{(2^n)^{2^n-1}}{(2^n)^{2^n}}\\
        & = \frac{1}{2^n}.
      \end{align*}

      It is quite surprising that instead of a upper bound
      on $\Pr[\MacForge_{\A, \tilde{\Pi}}(n) = 1]$
      depending on $\A$ (and reached for $\A$ super smart),
      it is actually independent of $\A$.
    \item
      Let's now suppose that we have an adversary $\A$
      that win with non-neglishible probability against a PRF $F$
      and show that under this assumption we can build
      a distinguisher $\D$ for $F$.

      $\D$ will simply take a function $g$ as input
      and run $\A$ using $g$ to create the tags.
      He has $g$ so he can see if $\A$ wins or lose.
      If $\A$ wins he outputs $1$ otherwise it outputs $0$.

      We know that if $g$ is a pseudo random function,
      $\Pr[\MacForge_{\A, \Pi_g} = 1] = \frac{1}{2^n}$
      and if $g$ is a PRF
      $\Pr[\MacForge_{\A, \Pi_g} = 1] = \eta(n)$
      where $\eta$ is nonnegligible.
      We have therefore
      \[
        |\Pr[\D^{F_k}(1^n) = 1] - \Pr[\D^{f}(1^n) = 1]|
        = \left|\eta(n) - \frac{1}{2^n}\right|
      \]
      which is non-negligible.

      % Not sure it works
%    \item
%      Another simpler solution is possible. Using the first Hint, we can say that if $F_k$ is a PRF, it has a maximum of $2^n$ possible outputs
%      where a truly random function has exactly $2^n$ outputs. So if we suppose $\Pi$ secure with a PRF, then $\tilde{\Pi}$ is also secure because
%      $\epsilon_{\tilde{\Pi}}  = \frac{1}{2^n} \geq \epsilon_{\Pi}$. We then can play the PRF game to prove the security of $\Pi$ with the second hint.
  \end{itemize}

\end{solution}

\subsection{}
\begin{solution}
  If we have the tag of $p$, which is (let's consider that $k$ and $p$ are $l$ bits long for simplicity)
  \[ t_p = H^s(k\|p) = h^s(h^s(h^s(h^s(IV \| k) \| p)) \| 2l) \]
  we can find the tag of $p\|2l\|w$ (where $w$ is $l$ bits long for simplicity)
  without knowing $k$ since we know $h^s$.
  It is
  \begin{align*}
    H^s(k\|p\|2l\|w)
    & = h^s(h^s(h^s(h^s(h^s(h^s(IV \| k) \| p)) \| 2l) \| w) \| 4l)\\
    & = h^s(h^s(H^s(k \| p) \| w) \| 4l)\\
    & = h^s(h^s(t_p \| w) \| 4l)
  \end{align*}
  Since $p\|2l\|w \neq p$, this gives us an existential forgery.
\end{solution}

\subsection{}
\begin{solution}
  From $L_0\|R_0$, we get $L_2\|R_2$
  where

  \begin{align*}
    L_2 & = L_0 \xor F_k(R_0)\\
    R_2 & = R_0 \xor F_k(L_0 \xor F_k(R_0)).
  \end{align*}
  When a distinguisher $\D$ sends $L_0\|R_0$ and receives
  $L_2\|R_2$ we can
  compute

  \begin{align*}
    L_2' & = L_0 \xor L_2\\
         & = L_0 \xor (L_0 \xor F_k(R_0))\\
         & = F_k(R_0)\\
    R_2' & = R_0 \xor R_2\\
         & = R_0 \xor (R_0 \xor F_k(L_0 \xor F_k(R_0)))\\
         & = F_k(L_0 \xor F_k(R_0)).
  \end{align*}
  The problem is that $L_2'$ does not depend on $L_0$.

  He can therefore send another request with the same $R_0$
  and a different $L_0$.
  If it gets the same $L_2'$,
  It has only 1 chance over $2^l$ (where $l$ is the number of bits of $L$) to receive the same $L_2'$ for a random permutation.

  If $\D$ decide to output 1 iff $L_2'$ is the same for the two request we have
  \[ |\Pr[\D^{F_k}(1^n) = 1] -\Pr[\D^f(1^n)]| = 1 - \frac{1}{2^n} \]
  which is clearly not negligible.
\end{solution}

\subsection{}
\begin{solution}
  Let say we have to split the message into $m$ messages of $n$ bits.
  Choosing $m_0 = M_0 \| M_0 \| \cdots \| M_0$ and $m_1 = M_1 \| M_2 \| \cdots \| M_m$ with $M_i \neq M_j$ for $i \neq j$,
  if $b = 0$, we will get $c = C_0 \| C_0 \| \cdots \| C_0$ for some $C_0 \in \C$ and $c = C_1 \| C_2 \| \cdots \| C_m$ for some $C_i \in \C$.

  An adversary $\A$ can output $b = 0$ iff all the $C_i$s are equals.
  We have
  \[ \Pr[\PrivKeav_{\A,\text{ECB}}(nm)] = \frac{1}{2} + \frac{1}{2} = 1 \]
  since the $C_i$ cannot be equal if $b = 1$ since we use a PRP.
  If two different $M_i$ were encrypted as same $C_i$, decryption wouldn't be possible.
\end{solution}

\subsection{}
\begin{solution}
  We will give 2 collisions for $f_1$ and $f_2$.

  We know that $E(k, \cdot)$ is surjective because it must be injective and $\M = \C$.
  Therefore the decryption exists for all $c \in \C$ ! We will use it for the second collision of $f_1$.

  For $f_1$, we have the 2 following collisions

  \begin{align*}
    f_1(D(E(y,x),y), E(y,x))
    & = E(E(y,x), D(E(y,x), y)) \xor E(y,x)\\
    & = y \xor E(y,x)\\
    & = E(y,x) \xor y\\
    & = f_1(x, y)\\
    f_1(D(0, E(y,x) \xor y), 0)
    & = E(0, D(0, E(y, x) \xor y)) \xor 0\\
    & = E(y, x) \xor y\\
    & = f_1(x, y).
  \end{align*}
  and for $f_2$ we have

  \begin{align*}
    f_2(x, E(x,x)) & = E(x,x) \xor E(x,x)\\
                   & = 0 & \forall x \in \{0,1\}^l\\
    f_2(y, E(x,x)) & = E(y,y) \xor E(x,x)\\
                   & = E(x,x) \xor E(y,y)\\
                   & = f_2(x, E(y,y)).
  \end{align*}
\end{solution}

\subsection{}
\begin{solution}
  \begin{itemize}
    \item
      Let's show that from a collision of $H_4$, we generate a collision for $H_2$
      which prove that $H_4$ is collision resistant since $H_2$ is so.
      Let's suppose that we have $x_1\|x_2 \neq y_1\|y_2$ are such that $H_4(x_1\|x_2) = H_4(y_1\|y_2)$.
      \begin{itemize}
        \item
          If $H_2(x_1) \| H_2(x_1 \xor x_2) \neq H_2(y_1) \| H_2(y_1 \xor y_2)$,
          we have a colision for $H_2$ since their image by $H_2$ is identical.
        \item
          If $H_2(x_1) \| H_2(x_1 \xor x_2) = H_2(y_1) \| H_2(y_1 \xor y_2)$,
          we have $H_2(x_1) = H_2(y_1)$ \emph{and} $H_2(x_1 \xor x_2) = H_2(y_1 \xor y_2)$.
          \begin{itemize}
            \item If $x_1 \neq y_1$, we have a collision for $x_2$ since $H_2(x_1) = H_2(y_1)$.
            \item If $x_1 = y_1$, then $x_2 \neq y_2$ since $x_1\|x_2 \neq y_1\|y_2$.
              Therefore $x_1 \xor x_2 \neq y_1 \xor y_2$ and we have collision on $H_2$.
          \end{itemize}
      \end{itemize}
    \item
      $H_6$ is not collision resistant since $H_6(x_1\|x_2\|x_3) = H_6((x_1 \xor w)\|(x_2 \xor w)\|(x_3 \xor w))$
      for all $w$.
      Indeed, since $\xor$ is associative and commutative,
      \begin{align*}
        & = H_6((x_1 \xor w)\|(x_2 \xor w)\|(x_3 \xor w))
        & = H_3(H2((x_1 \xor w) \xor (x_2 \xor w))\|H2((x_2 \xor w) \xor (x_3 \xor w))\|H2((x_3 \xor w) \xor (x_1 \xor w)))
        & = H_3(H2(x_1 \xor (w \xor w) \xor x_2)\|H2(x_2 \xor (w \xor w) \xor x_3)\|H2(x_3 \xor (w \xor w) \xor x_1)))
        & = H_3(H2(x_1 \xor x_2)\|H2(x_2 \xor x_3)\|H2(x_3 \xor x_1)))\\
        & = H_6(x_1\|x_2\|x_3).
      \end{align*}
  \end{itemize}
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item Encrypt all the DVDs with $K_\text{root}$ !
    \item At the beginning of the DVD, encrypt a random key $K$ twice (using $K_2$ and then $K_3$): $Enc_{K_2}(K)$ and $Enc_{K_3}(K)$.

      We then encrypt all the content $M$ of the DVD using $K$. (We suppose $K_1$ is $K_{\text{root}}$)
      $K_2$ and $K_3$ are the keys associated to the 2 childs of the root.
    \item
      For every node $i$ on the path from the root to $r$, we must add an encryption of $K$ with the key of its child that is not in the path.
      For example, if $r = 10$ and $n = 16$, we must include the bold keys,
      so we will include the encryption of $K_7$ in a DVD which is quite something.
      \begin{center}
        \Tree [.{$K_{\text{root}} = K_1$}
          [.{$\mathbf{K_2}$}
            [.{$K_4$}
              [.{$K_8$}
                [.{0} ]
                [.{1} ]
              ]
              [.{$K_9$}
                [.{2} ]
                [.{3} ]
              ]
            ]
            [.{$K_5$}
              [.{$K_{10}$}
                [.{4} ]
                [.{5} ]
              ]
              [.{$K_{11}$}
                [.{6} ]
                [.{7} ]
              ]
            ]
          ]
          [.{$K_3$}
            [.{$K_6$}
              [.{$\mathbf{K_{12}}$}
                [.{8} ]
                [.{9} ]
              ]
              [.{$K_{13}$}
                [.{10} ]
                [.{11} ]
              ]
            ]
            [.{$\mathbf{K_7}$}
              [.{$K_{14}$}
                [.{12} ]
                [.{13} ]
              ]
              [.{$K_{15}$}
                [.{14} ]
                [.{15} ]
              ]
            ]
          ]
        ]
      \end{center}
  \end{enumerate}
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item TODO
    \item $\A'$ should pick a random $r$ and send to its oracle the $d$ queries $r\|l\|i\|m_i$ for $i = 1, \ldots, d$.
      and output $\langle r, t_1, \ldots, t_d \rangle$ where $t_i$ is the answer of the oracle to its $i$th query.
    \item For $\Pi$ a forgery is $(m, \langle r, t_1, \ldots, t_d \rangle)$ where $m$ is not one of its previous query
      and $t_i = \Mac_k(r\|l\|i\|m_i)$ for $i = 1, \ldots, d$ where $l$ is the length of $m$ and $m = m_1\| \cdots \|m_d$.
      \begin{itemize}
        \item
          If none of its previous query has the same $r$ and $l$, $r\|l\|1\|m_1$ cannot be a query made by $\A'$ and
          $(r\|l\|1\|m_1, t_1)$ is an existential forgery for $\Pi'$ that $\A'$ can output.
        \item
          If 2 previous queries have both $r$ and $l$, then we do not necessarily have an existential forgery.
          However, since $r$ is picked at random ($\A'$ could cheat and make sure that the same $r$ is not picked twice)
          the probability (birthday paradox) of this to happen (if all $m$ have the same $l$ which is the worst case) is approximatively
          $\frac{q(n)^2}{2 \cdot 2^n}$ where $q(n)$ is the number of queries make by $\A$.
        \item
          If one unique previous query $m^j$ has this $r$ and $l$, since $m$ is not one of the previous query, there must be $i$
          such that $m_i \neq m_i^j$.
          We know therefore that $r\|l\|j\|m_j$ has never been queried by $\A'$ so $(r\|l\|j\|m_j, t_j)$ is an existential forgery for $\Pi$.
      \end{itemize}
    \item In conclusion, we have
      \begin{align*}
        \Pr[\MacForge_{\A,\Pi}(n) = 1]
        & \leq \Pr[\MacForge_{\A',\Pi'}(n) = 1] + \frac{q(n)}{2^{n+1}}\\
        \epsilon(n)
        & \leq \epsilon'(n) + \frac{q(n)}{2^{n+1}}
      \end{align*}
      so since $\epsilon'(n)$ and $\frac{q(n)}{2^{n+1}}$ are negligible,
      $\epsilon(n)$ is negligible.
  \end{enumerate}
\end{solution}

\subsection{}
\begin{solution}
  We build $H(m) = H_0(m)\|H_1(m)$.
  If we have $m_1 \neq m_2$ such that $H(m_1) = H(m_2)$ then $H_0(m_1) = H_0(m_2)$ and
  $H_1(m_1) = H_1(m_2)$ so the collision resistant hash function has a collision whichever it is.
  However, $H$ is no more a compression function and we cannot use Merkle-Damg\aa{}rd.

  The input of $H$ therefore cannot have arbitrary length but its input is twice the length of the output of $H_0$ and $H_1$
  so it is twice the size of a tag.

  The output of $\Mac_0$ and $\Mac_1$ are the size of a tag so we can use the tag
  $H(\Mac_0(k,m)\|\Mac_1(k,m))$ for our Hash-MAC scheme.
  If we are able to output an existential forgery $(m, t)$,
  since $H$ is collision resistant, that means that we have found
  $\Mac_0(k,m)\|\Mac_1(k,m)$ and therefore we have found an existantial forgery for both
  $\Mac_0$ \emph{and} $\Mac_1$ which is absurd since one of them is believed to be unforgeable.

  Our Hash-MAC scheme is therefore unforgeable.
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item When player $\ell$ receives $(i,c_0,c_1)$ he can compute its binary representation $\ell = \sum_{j=0}^{n-1} b_j2^j$
      and then computes
      \begin{align*}
        D(k_{i,b_i}, c_{b_i})
        & = D(k_{i,b_i}, E(k_{i,b_i}, m))\\
        & = m.
      \end{align*}
    \item The content provider can recover $b_j$ by feeding $P$ with $(i,E(k_{j,0},0),E(k_{j,1},1))$.
      He will then listen to what $P$ broadcast.
      $b_j$ will be what he will hear.
    \item $i \neq j$ and $i \xor j$ is not a power of two.
      That means that $i$ and $j$ differ for at least 2 bits.
      For one of them they will use the key of $i$ and for the other they will use the key of $j$.
      For example, if $i = 101_2$ and $j = 110_2$, they will use $k_{2,1}$, $k_{1,0}$ and $k_{0,0}$.
      That way the content provider will think that the index is $100_2$.
  \end{enumerate}
\end{solution}

\subsection{}
\begin{solution}
  Pour un message $m = m_1 \| m_2$,

  \[ \Enc_k(m_1, m_2) = (y_1, y_2, y_3)  =(p_k(m_1) , p_k(y_1 \xor m_2), pk(y_2 \xor r)) \]

  Le $R$ est random, et je pense que contrairement à ce que tu disais dans ta
  solution, il est fourni en texte clair en même temps que le cipher text
  (enfin bon de toute manière du point de vue de l'adversaire tu verras que
  ça ne change pas grand chose)

  Si on s'intéresse au ``game'' qui garantit la caractère CPA du scheme (cf slide 15 L3).
  Il suffit qu'un adversaire $\A$ choisisse $m_0$ et $m_1$ distinct, \emph{il peut demander
  leurs encryptions}.

  If $m_0 = m_W || m_X$ and $m_1 = m_Y || m_Z$

  \begin{align*}
    \Enc_k(m_W,m_X) = (p_k(mW), \ldots)\\
    \Enc_k(m_Y,m_Z) = (p_k(mY), \ldots).
  \end{align*}

  $\A$ envoit $m_0$ et $m_1$ au testeur
  Le testeur choisit $b$ et renvoie $Enc(m_b) = (y_1 ,y_2, y_3)$

  \begin{itemize}
    \item Si $y_1 = p_k(m_W)$ l'adversaire sait que $b = 0$
    \item Si $y_1 = p_k(m_Y)$ l'adversaire sait que $b = 1$
  \end{itemize}

  Le problème avec ce scheme d'encryption est qu'il n'est pas CPA secure car
  le facteur $r$ random est très mal appliqué, ce qui fait que le début du
  message encrypté qu'il renvoie est tout à fait déterministe et donc pas du
  tout CPA secure.
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item
      The problem with a MAC is that every party that can verify a MAC can also build it since verifying a MAC consist
      in recomputing the tag and comparing it the the received tag.
      All $B_i$ therefore need to have the key $k$ and can therefore send valid packets to other $B_j$.
    \item
      For $B_i$ to build all the MACs for $B_j$, he needs $S_i \subseteq S_j$.
      We therefore need that there is no $i \neq j$ such that $S_i \subseteq S_j$.
    \item We can simply take all $S_i$ having 2 elements.
      $S_i \subseteq S_j$ simplifies therefore in $S_i \neq S_j$.
      There are ${5 \choose 2} = \frac{5!}{2! \cdot 3!} = 10$ possible such sets.
      %Sure ? Because I think in this case B_k would be allowed to sign packets to other users haveing k_k in his set. No ?
  \end{enumerate}
\end{solution}
