\section{}
\subsection{}
\begin{solution}
  Two solution have been proposed
  \begin{itemize}
    \item
      Let's show that if $f$ is not one way, then $\Pi$ is not existentially unforgeable.
      Let $\A$ be the inverter of $f$, we will build $\A'$ that builds an existential forgery with non-negligible probability.

      \begin{itemize}
        \item $\A'$ receives $pk = (y_0, y_1)$
        \item $\A'$ ask the signature of 0 and gets $\sigma$, he does not really care about it
        \item $\A'$ gives $y_1$ to $\A$ which outputs $x_1'$
        \item $\A'$ outputs $(1, x_1')$
      \end{itemize}
      Since $y_1$ is the image of a random $x_1$, we are exactly in the inverting experiment so $f(x_1') = y_1$
      whith probability $\Pr[\Invert_{\A,f}(n) = 1]$.

      We know that
      \[
        \Pr[\Sigforgeone_{\A',\Pi}(n) = 1] = \Pr[\Invert_{\A,f}(n) = 1]
      \]
    \item
      Let assume that $f$ is not one way function $y = f(x)$.
      Then, we can recover $x$ with a non negligible probability $\epsilon_x (n)$.

      So, $(y_0, y_1) \Rightarrow (x_0, x_1)$ with probabilities $(\epsilon_{x_0} (n), \epsilon_{x_1} (n))$.
      We cannot compute a pre-image by asking the oracle. So I output $(0, x_1)$ and $(1, x_2)$ as a forgery.

      The probability $\Pr[\Sigforge(y)] = 2\epsilon(n)$, non-negligible since f is not one way.
  \end{itemize}
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item
      A has $(i, \sigma(i))$ whith $\sigma (i) = f^{n-i} (x)$. We know (because $f$ is a permutation function) that:
      $$f(\sigma(i)) = f^{n-i+1}(x) = f^{n-(i-1)}(x) = \sigma(i-1)(x)$$
      Then it's possible to compute a valid forgery for every $j < i$. The scheme is then not a one time-signature.
    \item
      Need schema drawn at TP !! It's a lot simpler with it...

      $\Pr[Success (A_\sigma)] = \epsilon(\lambda))$, $\Pr[Abort] = \frac{n-k}{n}$, $\Pr[Success] = \frac{n-k}{n-m-1}$ then:
      $$ \Pr[Success(A_{owf})] = \epsilon(\lambda) \frac{n-k}{n} \frac{n-k}{n-m-1}$$

      If $\epsilon(\lambda)$ is not negligible, then the probability of success is not negligible.
    \item
      We have $s_k = (x, x')$, $p_k = (f^n(x), f^n(x'))$.
      Then $m \rightarrow \sigma = (f^{n-m}(x), f^m(x'))$.

  \end{enumerate}
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item
      $(r, s) = ([g^k \pmod{p}] \pmod{q}, [H(m) + xr]k^{-1} \pmod{q})$, then:
      $u_1 = H(m)s^{-1} \pmod{q}$,  $u_2 = rs^{-1} \pmod{q}$, $r = [g^{u_1} g^{u_2} \pmod{p}] \pmod{q}$, $y = g^x$

      $$\Rightarrow g^{u_1 + xu_2} = g^{s^{-1}(H(m) + rx)} = [g^k \pmod{p}] \pmod{q} = r$$
    \item
      $s = (H(m) + xr)k^{-1} \pmod{q}$, $s' = (H(m') + xr)k^{-1} \pmod{q}$ ($s \neq s'$ otherwise we have a collision).
      $s - s' = (H(m) - H(m'))k^{-1} \pmod{q}$, $k = \frac{H(m) - H(m')}{s - s'}$.
      $s = (H(m) + xr)k^{-1}$ so $\frac{sk - H(m)}{r} = x$ where x is the secret.
  \end{enumerate}
\end{solution}

\subsection{}
\begin{solution}
  \begin{enumerate}
    \item
      It is not a secure scheme.
      We can ask for the signature of $m_1$ which is $(\sigma_1, \sigma_2, s)$ and give the existential forgery
      $(m, (\sigma_1, \Mac_s(m), s))$.
      The signature $\sigma_1$ has been replayed
    \item
      It is a secure scheme.
      Let's prove that if $\A$ an produce an existential forgery,
      we can build one for either the regular signature scheme or the one-time signature scheme.

      $\A$ does the $q(n)$ queries $m_1, \ldots, m_{q(n)}$ and receive the signatures $(\sigma_{1;i}, \sigma_{2;i}, {vk'}_i)$.
      If it then outputs an existential forgery $(m, (\sigma_1, \sigma_2, vk'))$, let's consider 2 cases
      \begin{itemize}
        \item
          Either $vk' \neq {vk'}_i$ for all $i$.
          Does that mean that $\sigma_2$ is an existential forgery for the one-time signature scheme ?
          We could say ``\emph{Not necessarily since $sk',vk'$ are picked at random, with very bad luck, two same could have been picked.
          And even if they are different, thinking about Lamport, some $y$ could be the same between two keys !}''.
          However here we deal with \emph{any} one-time signing scheme so we cannot reason about Lamport at a lower level and find an invertion of $f$.

          Fortunately, this is absolutely not a problem.
          An adversary $\A'$ trying to find an existential forgery for the one-time signature scheme has access $\Sig$,
          it is allowed to pick new random $(sk', vk')$ and compute $\sigma_2$ for those.
          Since it is PPT, it only has the time to pick a polynomial number of those so it is not a problem.

          Let's do a reduction where it is clear that there is no problem
          \begin{enumerate}
            \item $\A'$ receives $vk'$ and pick a random $i^* \in \{0, \ldots, q\}$ (he tries to guess for which ${vk'}_i$ will be reused).
            \item $\A'$ gives $pk$ to $\A$
              \begin{itemize}
                \item When $\A$ sends $m_i$ for $i \neq i^*$, $\A'$ pick random $({sk'}_i, {vk'}_i)$ and
                  computes sends $(\Sign_{sk}({vk'}_i), \Sig_{{sk'}_i}(m_i), {vk'}_i)$ to $\A$.
                \item When $\A$ sends $m_i$ for $i = i^*$, $\A'$ asks its single query $m_i$ and receives $\sigma_{2;i} = \Sig_{sk'}(m_i)$.
                  He computes $\sigma_{1;i} = \Sign_{sk}({vk'}_i)$ and sends $(\sigma_{1;i}, \sigma_{2;i}, vk')$ to $\A$.
              \end{itemize}
            \item $\A$ tries outputs an existential forgery, if it is valid and $vk' = {vk'}_{i^*}$ (probability $1/q(n)$),
              $\A'$ also has an existential forgery and wins.
          \end{enumerate}
        \item
          Either $vk' \neq {vk'}_i$ for all $i$.
          That means that $\sigma_1$ is an existential forgery for the regular signature scheme.

          From an adversary $\A$ for this scheme, we
          can build an adversary $\A''$ for the regular signature scheme
          simply outputs $(m,\sigma_1)$ when $\A$ outputs its
          existantial forgery $(\sigma_1, \sigma_2, m)$
          hoping that $vk'$ is not the same as an previous ${vk'}_i$.
          In that case, it will win if $\A$ wins.
      \end{itemize}

      In summary, we have
      \begin{align*}
        \Pr[\Sigforge_{\A}(n) = 1]
        & = \Pr[\Sigforge_{\A}(n) = 1 \land (\exists i: vk' = {vk'}_i)]\\
        & \quad + \Pr[\Sigforge_{\A}(n) = 1 \land (\forall i: vk' \neq {vk'}_i)]\\
        & \leq \Pr[\Sigforge_{\A''}(n) = 1] + \frac{\Pr[\Sigforgeone_{\A'}(n) = 1]}{q(n)}
      \end{align*}
      which is non-negligible since $\Pr[\Sigforge_{\A''}(n) = 1]$ and $\Pr[\Sigforgeone_{\A'}(n) = 1]$ are negligible and $q(n)$ is polynomial.
  \end{enumerate}
\end{solution}

\subsection{}
\copypaste{4}{1}

\subsection{}
\copypaste{4}{2}

\subsection{}
\copypaste{4}{3}

\subsection{}
\copypaste{4}{4}

\subsection{}
\copypaste{4}{5}
