
\subsection{Impact}
Compared to other technologies (video, GSM, \ldots), RFID has several issue
concerning privacy:
\begin{itemize}
    \item Tags cannot be switched-off
    \item Passive tags answer without the agreement of their bearers
    \item Easy to analyze the logs of the readers
    \item Increasing of the communication range
    \item Tags can be almost invisible
\end{itemize}

Privacy issues concerning RFID is being seriously considered by authorites.

\subsection{Classification}

\begin{itemize}
    \item \textbf{Information Meaningful by Itself}: Privacy issues
        appear when the data sent by the tag reveals information
        intrinsic to the marked object or the holder of the object.

        \paragraph{Public transportation}
        MOBIB card are rather powerful RFID tags that embed
        cryptographic mechanisms to avoid impersonation or cloning,
        Personal data are stored in the clear in the card.
        %TODO understand

    \item \textbf{Information Meaningful with the database}: There exist
        website (or technique in general) to make the link between
        anonymous information in database.
\end{itemize}

\subsection{Mitigating the Problem}
\begin{itemize}
    \item Kill command to destroy the tag
    \item Use a faraday cages or a blocker tags to
        avoid clandestine query.
    \item Remove antenna
    \item $\cdots$
\end{itemize}
The problem with these solutions is that they are not convenient.

\subsubsection{Privacy problem}

Nowadays more and more data are collected, it is called
logphilia. But with all these data, information may eventually leak.

The consequence of these information's leakage must be evalutated:
\begin{itemize}
    \item Does all these data need to be store?
    \item Encryption of the sensitive data on the tag in the databases
    \item Authentication for accessing the data
    \item Establishing policies to define who can access the data
\end{itemize}

\subsubsection{Future}
\begin{itemize}
    \item Cryptographic building blocks available in the tags are more
        secure in recent products (lightweight implementation of
        standardized algorithms)
    \item Note that Secure building blocks do not make themselves secure
        applications: 
        \begin{itemize}
            \item The security of the whole application must be considered.
            \item Many SMEs involved in RFID.
        \end{itemize}
\end{itemize}

