\subsection{} % Exercise 1.
The following Turing machine has the alphabets \(\Sigma = \{a, b\}\)
and \(\Gamma = \{a, b, x, B, 0, 1\}\).
What does it compute?

\begin{center}
\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $a$ & seekB & $\to$ & $x$ \\
	\hline
	start & $b$ & seekA & $\to$ & $x$ \\
	\hline
	start & $x$ & start & $\to$ & $B$ \\
	\hline
	start & $B$ & stop & $\downarrow$ & $1$ \\
	\hline
	seekA & $a$ & restart & $\leftarrow$ & $x$ \\
	\hline
	seekA & $b$ & seekA & $\to$ & $b$ \\
	\hline
	seekA & $x$ & seekA & $\to$ & $x$ \\
	\hline
	seekA & $B$ & false & $\leftarrow$ & $B$ \\
	\hline
	seekB & $b$ & restart & $\leftarrow$ & $x$ \\
	\hline
	seekB & $a$ & seekB & $\to$ & $a$ \\
	\hline
	seekB & $x$ & seekB & $\to$ & $x$ \\
	\hline
	seekB & $B$ & false & $\leftarrow$ & $B$ \\
	\hline
	restart & $a$ & restart & $\leftarrow$ & $a$ \\
	\hline
	restart & $b$ & restart & $\leftarrow$ & $b$ \\
	\hline
	restart & $x$ & restart & $\leftarrow$ & $x$ \\
	\hline
	restart & $B$ & start & $\to$ & $B$ \\
	\hline
	false & $a$ & false & $\leftarrow$ & $B$ \\
	\hline
	false & $b$ & false & $\leftarrow$ & $B$ \\
	\hline
	false & $x$ & false & $\leftarrow$ & $B$ \\
	\hline
	false & $B$ & stop & $\downarrow$ & $0$ \\
	\hline
\end{tabular}
\end{center}
\noindent \(\downarrow\) represents the stay of the head.
\bigskip

Hints:
\begin{itemize}
	\item What is the result for input ``\(bbba\)''?
	And for ``\(baba\)''?
	\item What are the two first cells changed to \(x\)
	if the string starts with an ``\(a\)''?
	And a ``\(b\)''?
	\item How are the \(x\)'s handled in the start state?
	\item What does an \(x\) mean?
	\item What is the purpose of each state?
\end{itemize}

\begin{solution}
	\begin{itemize}
		\item The result for ``\(bbba\)'' is \(0\),
		while the result for ``\(baba\)'' is \(1\).
		\item If the string starts with ``\(a\)'',
		the first two cells changed to \(x\) are that first ``\(a\)''
		and the first ``\(b\)'' to its right.
		Conversely, if the string starts with ``\(b\)'',
		the first two cells changed to \(x\) are that first ``\(b\)''
		and the first ``\(a\)'' to its right.
		\item In the start state, ``\(x\)''
		are erased and replaced by blank characters.
		\item An ``\(x\)'' denotes a placeholder character
		for when the character we are looking for has been found.
		At the end of the execution, all ``\(x\)''s
		are erased and replaced by blank characters.
		\item The ``start'' state launches the program.
		The ``seekA'' and ``seekB'' states
		respectively look for ``\(a\)'' and ``\(b\)'',
		replace them with ``\(x\)'' when found and then restart.
		The ``restart'' state gets the program
		back to the beginning of the tape and in the ``start'' state.
		The ``false'' state clears the tape and puts a single ``\(0\)''.
		The ``stop'' state halts execution,
		either returning \(0\) or \(1\).
	\end{itemize}

	This program determines whether the number of ``\(a\)''s and ``\(b\)''s
	in the input string is the same.
\end{solution}

\subsection{} % Exercise 2.
For each of the following instructions, build a Turing machine,
using \(\Sigma = \{0, 1\}\):
\begin{enumerate}
	\item \(f(n) = n \times 2\);
	\item \(f(n) = \mathrm{not}(n)\) (\(f\) invers the bits of \(n\)).
\end{enumerate}

\begin{solution}
\begin{enumerate}
	\item Doubling a number in binary is the same
	as adding a \(0\) at its end.
	This means that a valid Turing machine would be
	\begin{center}
	\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $1$ & start & $\to$ & $1$ \\
	\hline
	start & $0$ & start & $\to$ & $0$ \\
	\hline
	start & $B$ & return & $\leftarrow$ & $0$ \\
	\hline
	return & $1$ & return & $\leftarrow$ & $1$ \\
	\hline
	return & $0$ & return & $\leftarrow$ & $0$ \\
	\hline
	return & $B$ & stop & $\to$ & $B$ \\
	\hline
	\end{tabular}
	\end{center}
	This Turing machine first goes from left to right, places an extra \(0\)
	and then goes back to the start of the number.
	\item A Turing machine which does this is the following:
	\begin{center}
	\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $1$ & start & $\to$ & $0$ \\
	\hline
	start & $0$ & start & $\to$ & $1$ \\
	\hline
	start & $B$ & return & $\leftarrow$ & $B$ \\
	\hline
	return & $1$ & return & $\leftarrow$ & $1$ \\
	\hline
	return & $0$ & return & $\leftarrow$ & $0$ \\
	\hline
	return & $B$ & stop & $\to$ & $B$ \\
	\hline
	\end{tabular}
	\end{center}
	This Turing machine first goes from left to right, negates every bit
	and then goes back to the start of the number.
\end{enumerate}
\end{solution}

\subsection{} % Exercise 3.
Suppose we have two Turing machines \(A\) and \(B\)
with \(\Sigma = \{a, b\}\).
Their starting and halting states
are \((s_A, h_A)\) and \((s_B, h_B)\), respectively.
Explain how to combine \(A\) and \(B\) into a new Turing machine implementing
the following pseudo-code:
\begin{enumerate}
	\item \texttt{if($a$) then {A} else {B}}
	\item \texttt{while not($a$) do {A}}
\end{enumerate}
In this pseudocode, \texttt{if($a$)} means
``if $a$ is the symbol under the head'',
and \texttt{not($a$)} means ``the symbol under the head is not $a$'',
none of which consumes a symbol.

\begin{solution}
\begin{enumerate}
	\item We can give the following Turing machine:
	\begin{center}
	\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $a$ & $s_A$ & $\downarrow$ & $a$ \\
	\hline
	start & $b$ & $s_B$ & $\downarrow$ & $b$ \\
	\hline
	start & $B$ & $s_B$ & $\downarrow$ & $B$ \\
	\hline
	$h_A$ & $a$ & stop & $\downarrow$ & $a$ \\
	\hline
	$h_A$ & $b$ & stop & $\downarrow$ & $b$ \\
	\hline
	$h_A$ & $B$ & stop & $\downarrow$ & $B$ \\
	\hline
	$h_B$ & $a$ & stop & $\downarrow$ & $a$ \\
	\hline
	$h_B$ & $b$ & stop & $\downarrow$ & $b$ \\
	\hline
	$h_B$ & $B$ & stop & $\downarrow$ & $B$ \\
	\hline
	\end{tabular}
	\end{center}
	\item We can give the following Turing machine:
	\begin{center}
	\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $a$ & stop & $\downarrow$ & $a$ \\
	\hline
	start & $b$ & $s_A$ & $\downarrow$ & $b$ \\
	\hline
	start & $B$ & $s_A$ & $\downarrow$ & $B$ \\
	\hline
	$h_A$ & $a$ & stop & $\downarrow$ & $a$ \\
	\hline
	$h_A$ & $b$ & start & $\downarrow$ & $b$ \\
	\hline
	$h_A$ & $B$ & start & $\downarrow$ & $B$ \\
	\hline
	\end{tabular}
	\end{center}
\end{enumerate}
\end{solution}

\subsection{} % Exercise 4.
Are these Turing machines equivalent to the standard model of Turing machine?
Explain.
\begin{enumerate}
	\item ``Big jumps'': this Turing machine
	can move its head of \(n\) cells to the left and to the right.
	The transition function is thus
	\(S \times \Gamma \to S \times (\{L, R\} \times \N) \times \Gamma\).
	\item ``Online'': this Turing machine can only move to the right
	or stay at the same cell,
	and cannot move to the left.
	The transition function is thus
	\(S \times \Gamma \to S \times \{R, \downarrow\} \times \Gamma\).
\end{enumerate}

\begin{solution}
	\begin{enumerate}
		\item This Turing machine is equivalent
		to the standard Turing machine.
		Since \(n\) small steps can now be grouped
		into one big step of size \(n\),
		the model has the same power,
		while offering a linear speedup
		over the standard Turing machine.
		\item This Turing machine is not equivalent
		to the standard Turing machine.
		Consider for example a standard Turing machine
		which inverts the first bit of input
		if the last bit is \(0\).
		This machine cannot be translated to an online Turing machine.
	\end{enumerate}
\end{solution}

\subsection{} % Exercise 5.
True or false?
\begin{enumerate}
	\item If a Turing machine decides a language \(L\),
	then \(L\) is recursive.
	\item A Turing machine execution always terminates.
	\item ND-Turing machines are
	a more expressive formalism than Turing machines.
	\item Turing machines with oracle for
	\(A = \{x \in \N \mid x \% 2 = 0\}\) are
	more expressive than Turing machines.
	\item Turing machines with oracle for
	\(A = \{x \in \N \mid \phi_x(x) = 42\}\) are
	more expressive than Turing machines.
\end{enumerate}

\begin{solution}
\begin{enumerate}
	\item True.
	Turing machines are a valid model of computability,
	hence if one of them computes the characteristic function for \(L\),
	then this function is computable,
	and the language is recursive.
	\item False.
	Consider the following Turing machine, on \(\Sigma = \emptyset\):
	\begin{center}
	\begin{tabular}{|c|c||c|c|c|}
	\hline
	\textbf{State} & \textbf{Symbol} & \textbf{State} & \textbf{Movement} & \textbf{Symbol} \\
	\hline
	start & $B$ & start & $\to$ & $B$ \\
	\hline
	\end{tabular}
	\end{center}
	This machine computes \(\bot\).
	\item False.
	An ND-Turing machine has the same expressive power
	as a standard Turing machine, but achieves an exponential speedup.
	It is however not possible to build such a machine,
	since one has to simulate the nondeterminism.
	\item False.
	Since this set is recursive,
	this Turing machine has the same expressive power
	as a standard turing machine.
	\item True.
	This set is not recursive.
	To prove it, assume it is recursive.
	We build a program for fixed \(n, k\)
	which we assume to have Gödel number \(d\):
\begin{minted}{julia}
function P_d(q)
	P_n(k)
	return 42
end
\end{minted}
	Since we made the assumption that \(A\) is recursive,
	we can construct a program that decides \(\HALT\) as follows:
\begin{minted}{julia}
function P_H(n, x)
	Build P_d(q) # d depends on n and k.
	return P_A(d)
end
\end{minted}
	However, we know that this program cannot exist.
	This means that our initial assumption is wrong, that is,
	\(A\) is not recursive.
	Since it is not recursive,
	the Turing machine with oracle
	is more expressive than a standard Turing machine.
\end{enumerate}
\end{solution}

\subsection{} % Exercise 6.
What is the computational complexity of the Turing machine
described in Exercise~6.1?
Let TM2T be a Turing machine able to use two strips of tapes instead of one.
Show that there exists a TM2T that computes
the same function as the TM of Exercise~6.1,
while achieving a quadratic computational speedup.

\begin{solution}
	The worst case input for this Turing machine
	is a string of the form \(\overbrace{a \cdots\vphantom{b} a}^{n \textnormal{ times}}\overbrace{b \cdots b}^{n \textnormal{ times}}\)
	(or the opposite, of course),
	In this case, the Turing machine will traverse \(n\) characters,
	looking for a ``\(b\)'',
	then do \(n\) steps back to restart its execution.
	Then it will do \(n+1\) steps, again looking for a ``\(b\)''.
	It will do this \(n\) times,
	and finally will go back from right to left,
	until it reaches a blank character, needing another \(2n\) steps.
	This in total equals around \(\bigoh(n^2)\) steps,
	meaning the Turing machine has quadratic complexity.

	We can easily construct this Turing machine:
	consider that the first tape contains the input string,
	while the second tape starts out blank.
	The first tape simply reads from left to right,
	and writes its results on the second tape,
	tallying the number of ``\(a\)``s and ``\(b\)''s,
	for example by copying the character read by the first tape
	if the tape contains no other characters,
	or by rewriting one of the other characters
	to be a blank character if it does.
	At the end of the execution, simply check whether the second tape's head
	is reading a blank character and decide accordingly:
	if yes, then the number of ``\(a\)''s
	is equal to the number of ``\(b\)''s,
	if no, then not.
	This multitape Turing machine has linear complexity
	(the first head passes over the input characters once
	before the Turing machine halts),
	whereas its standard counterpart has quadratic complexity.
	This shows that the improvement is quadratic.

	In fact, for any number of tapes and for any multitape Turing machine
	with complexity \(\bigoh(t(n))\)
	there exists an equivalent single-tape standard Turing machine
	which has complexity \(\bigoh(t(n)^2)\).
\end{solution}
