\clearpage{}
\section{Define reliability, maintainability and availability, and their
expression as MTBF and MTTR.\@ Discuss the case of a constant failure rate.
Explain the principles of failure prediction models. Explain and discuss
the principles of statistical testing.}


\subsection{RAM attribute (reliability, availability, maintainability)}

\begin{description}
    \item[Reliability] is the probability that a system will perform its intended function satisfactorily for a given 
    time interval. R(t) = Prob(no failure before t)
    \item[Maintainability] is the probability that maintenance of the system will retain the system in, or restore it 
    to, a specified condition within a given time period. M(t) = Prob(restored before t)
    \item[Availability] is the probability that the system is operating satisfactorily at any time, and it depends on the reliability and the maintainability.  A = Prob(not failed)
\end{description}

Hence the study of probability theory is essential for understanding the reliability, maintainability, and availability of the system.

\subsection{Mean time between failures (MTBF)}
Average time between successive failures. Measures reliability.

\subsection{Mean time to repair (MTTR)}
Average time to fix a fault. Measures maintainability. \newline
Availability: A = MTBF / (MTBF + MTTR)

\subsection{Constant failure rate}
The system will fail at constant probability rate of once per hour. \newline

Then, MTBF = $3600 s$ \newline
Density function: $f(t) = 1/3600 * \exp( -t / 3600)$ (exponential law) \newline
Distribution function: $F(t) = 1 - \exp(-t / 3600)$ \newline
Reliability $(R (t) = 1 - F(t)) : R(t) = \exp(-t / 3600)$ \newline
Reliability $(\textrm{for t} = MTBF): R(MTBF) = \exp(-1) = 37\%$ \newline

\subsection{Failure prediction models}
Goal: Given a series of failure times ${t}_{1}, {t}_{2},\ldots, {t}_{n-1}$ \newline
Predict the distribution of future failure times ${T}_{n}, {T}_{n+1},\ldots$ \newline

A good prediction model must include:

\begin{itemize}
    \item Prediction model: probability specification of the stochastic process.
    \item Inference procedure: infer unknown parameter from ${t}_{1}\ldots{t}_{n-1}$.
    \item Prediction procedures: combine model and inference procedure to make predictions about future failure behaviour.
\end{itemize}

\subsection{Principles of statistical testing}

Reliability prediction based on failures occurring during testing \newline
Will this be accurate for typical system usage? \newline

\begin{description}
    \item[Operational profile] probability distribution on inputs (reflecting usage)
    \item[Statistical testing] select tests according to operational profile
        \subitem{} Tests focus on more used parts ($\Rightarrow$ better observable reliability)
        \subitem{} Test reflect usage ($\Rightarrow$ reliability predictions more accurate)
\end{description}

But operational profiles are difficult to define.\newline
Mislead, a small \% of operational profile may account for a large \% of failures.
