\section{Reliable broadcast}

\subsection{Best-effort broadcast}

\begin{itemize}
	\item Best-effort-Validity: if $p_i$ and $p_j$ are correct,
	then any broadcast by $p_i$ is eventually delivered by $p_j$
	\item No duplication: No message delivered  more than once
	\item No creation: No message delivered unless broadcast.
\end{itemize}

BeB gives no guarantee if the sender crash.

\input{listings/best_effort_broadcast.tex}

Implementation is straightforward, simple loop on all nodes.

\subsection{Reliable Broadcast}
\begin{itemize}
	\item Validity: same as BeB
	\item No duplication.
	\item No creation.
	\item Agreement: if a correct node delivers m, then every correct
	node delivers m.
\end{itemize}
\subsubsection{Uniform Reliable Broadcast}
If a failed node delivers, everyone must deliver.
It adds the \textbf{Uniform agreement}: for any message $m$,
if a process delivers $m$, then every correct process delivers $m$.

\subsubsection{Implementation}
\textbf{Using Beb and P}

    \input{listings/lazy_reliable_broadcast.tex}

\paragraph{Proof correctness}
If correct $p_j$ delivers msg broadcast by $p_i$
\begin{itemize}
	\item If $p_i$ is correct, BeB ensures correct delivery
	\item If $p_i$ crashes,
		\begin{itemize}
			\item $p_j$ detects this (completeness).
			\item $p_j$ uses BeB to ensure every correct node gets it.
		\end{itemize}
\end{itemize}
If we use a eventually perfect detector, it only affects
performance, not correctness.


\subsection{Eager Reliable Broadcast}
The exists a eager version of reliable broadcast that does not use a
perfect detector. It simply use a BeB broadcast upon receive of a
message.

    \input{listings/eager_reliable_broadcast.tex}

\subsubsection{Correctness of Eager RB}
if correct $p_j$ delivers message broadcasted by $p_i$,
$p_j$ uses BeB to ensure every correct node gets it.


\subsection{Uniform reliable broadcast (URB)}
In the above version, uniformity is not respected. If a sender
p immediately RB delivers and crashes, only p will have delivered
the message.

\input{listings/uniformity_reliable_broadcast.tex}

The idea is the following, message are pending until all correct
nodes get it. Node that gets the message exchange ack. The
message is deliver once all the nodes acked.

\subsubsection{Correctness}
\begin{itemize}
    \item No creation from BEB
    \item No duplication by using delivered set
\end{itemize}

\paragraph{Lemma} If a correct node $p_i$ bebDelivers $m$, then $p_i$ eventually
urbDelivers $m$.

\paragraph{Proof} Correct node $p_i$ bebBroadcasts $m$ as soon as it gets $m$

\begin{itemize}
    \item by BEB1 every correct node gets $m$ and bebBroadcasts $m$;
    \item $p_i$ gets $bebDeliver(m)$ from every correct node by BEB1;
    \item by completeness of $P$, it will not wait for dead nodes forever.
\end{itemize}

\paragraph{Majority-ACK Uniform Reliable Broadcast} If a majority of nodes
ack $m$, it reaches the agreement. We change $canDeliver(m)$ to return true when
$\mid ack[m] \mid > n/2$
