% These solutions come from the year 2014-2015 and do not have written statements
%% TODO find and copy the statements

%TP2 EX3
\begin{solution}
	From $L_0\|R_0$, we get $L_2\|R_2$
	where
	
	\begin{align*}
		L_2 & = L_0 \xor F_k(R_0)\\
		R_2 & = R_0 \xor F_k(L_0 \xor F_k(R_0)).
	\end{align*}
	When a distinguisher $\D$ sends $L_0\|R_0$ and receives
	$L_2\|R_2$ we can
	compute
	
	\begin{align*}
		L_2' & = L_0 \xor L_2\\
		& = L_0 \xor (L_0 \xor F_k(R_0))\\
		& = F_k(R_0)\\
		R_2' & = R_0 \xor R_2\\
		& = R_0 \xor (R_0 \xor F_k(L_0 \xor F_k(R_0)))\\
		& = F_k(L_0 \xor F_k(R_0)).
	\end{align*}
	The problem is that $L_2'$ does not depend on $L_0$.
	
	He can therefore send another request with the same $R_0$
	and a different $L_0$.
	If it gets the same $L_2'$,
	It has only 1 chance over $2^l$ (where $l$ is the number of bits of $L$) to receive the same $L_2'$ for a random permutation.
	
	If $\D$ decide to output 1 iff $L_2'$ is the same for the two request we have
	\[ |\Pr[\D^{F_k}(1^n) = 1] -\Pr[\D^f(1^n)]| = 1 - \frac{1}{2^n} \]
	which is clearly not negligible.
\end{solution}

%TP2 EX10
\subsection{}
\begin{solution}
	\begin{enumerate}
		\item When player $\ell$ receives $(i,c_0,c_1)$ he can compute its binary representation $\ell = \sum_{j=0}^{n-1} b_j2^j$
		and then computes
		\begin{align*}
			D(k_{i,b_i}, c_{b_i})
			& = D(k_{i,b_i}, E(k_{i,b_i}, m))\\
			& = m.
		\end{align*}
		\item The content provider can recover $b_j$ by feeding $P$ with $(i,E(k_{j,0},0),E(k_{j,1},1))$.
		He will then listen to what $P$ broadcast.
		$b_j$ will be what he will hear.
		\item $i \neq j$ and $i \xor j$ is not a power of two.
		That means that $i$ and $j$ differ for at least 2 bits.
		For one of them they will use the key of $i$ and for the other they will use the key of $j$.
		For example, if $i = 101_2$ and $j = 110_2$, they will use $k_{2,1}$, $k_{1,0}$ and $k_{0,0}$.
		That way the content provider will think that the index is $100_2$.
	\end{enumerate}
\end{solution}

% TP2 Ex11
\begin{solution}
	Pour un message $m = m_1 \| m_2$,
	
	\[ \Enc_k(m_1, m_2) = (y_1, y_2, y_3)  =(p_k(m_1) , p_k(y_1 \xor m_2), pk(y_2 \xor r)) \]
	
	Le $R$ est random, et je pense que contrairement à ce que tu disais dans ta
	solution, il est fourni en texte clair en même temps que le cipher text
	(enfin bon de toute manière du point de vue de l'adversaire tu verras que
	ça ne change pas grand chose)
	
	Si on s'intéresse au ``game'' qui garantit la caractère CPA du scheme (cf slide 15 L3).
	Il suffit qu'un adversaire $\A$ choisisse $m_0$ et $m_1$ distinct, \emph{il peut demander
		leurs encryptions}.
	
	If $m_0 = m_W || m_X$ and $m_1 = m_Y || m_Z$
	
	\begin{align*}
		\Enc_k(m_W,m_X) = (p_k(mW), \ldots)\\
		\Enc_k(m_Y,m_Z) = (p_k(mY), \ldots).
	\end{align*}
	
	$\A$ envoit $m_0$ et $m_1$ au testeur
	Le testeur choisit $b$ et renvoie $Enc(m_b) = (y_1 ,y_2, y_3)$
	
	\begin{itemize}
		\item Si $y_1 = p_k(m_W)$ l'adversaire sait que $b = 0$
		\item Si $y_1 = p_k(m_Y)$ l'adversaire sait que $b = 1$
	\end{itemize}
	
	Le problème avec ce scheme d'encryption est qu'il n'est pas CPA secure car
	le facteur $r$ random est très mal appliqué, ce qui fait que le début du
	message encrypté qu'il renvoie est tout à fait déterministe et donc pas du
	tout CPA secure.
\end{solution}

%TP2 Ex12
\begin{solution}
	\begin{enumerate}
		\item
		The problem with a MAC is that every party that can verify a MAC can also build it since verifying a MAC consist
		in recomputing the tag and comparing it the the received tag.
		All $B_i$ therefore need to have the key $k$ and can therefore send valid packets to other $B_j$.
		\item
		For $B_i$ to build all the MACs for $B_j$, he needs $S_i \subseteq S_j$.
		We therefore need that there is no $i \neq j$ such that $S_i \subseteq S_j$.
		\item We can simply take all $S_i$ having 2 elements.
		$S_i \subseteq S_j$ simplifies therefore in $S_i \neq S_j$.
		There are ${5 \choose 2} = \frac{5!}{2! \cdot 3!} = 10$ possible such sets.
		%Sure ? Because I think in this case B_k would be allowed to sign packets to other users haveing k_k in his set. No ?
	\end{enumerate}
\end{solution}

% TP3 Ex1
\begin{solution}
	\begin{enumerate}
		\item
		To have $m^{ed} \equiv m \pmod{N}$,
		it is sufficient to have $ed \equiv 1 \pmod{\varphi(N)}$.
		To have this, we need $e$ relatively prime to $\varphi(N)$.
		
		$\varphi(N) = (7-1)(11-1) = 60 = 2^235$
		so the smallest $e$ we can take is $7$.
		\item
		To find $d$ such that $ed \equiv 1 \pmod{60}$ we
		can use the extended Euclidean algorithm \footnote{A good explaination can be found at \url{https://www.youtube.com/watch?v=fz1vxq5ts5I}}
		
		\begin{align*}
			60 & = 8 \cdot 7 + 4\\
			7 & = 1 \cdot 4 + 3\\
			4 & = 1 \cdot 3 + 1
		\end{align*}
		so
		
		\begin{align*}
			1 & = 4 - 1 \cdot 3\\
			& = 2 \cdot 4 - 1 \cdot 7\\
			& = 2 \cdot 60 - 17 \cdot 7.
		\end{align*}
		We can take $-17 \equiv 43 \pmod{60}$ for $d$.
		\item
		We have
		
		\begin{align*}
			c & \equiv 75^7 \pmod{77}\\
			& \equiv (-2)^7 \pmod{77}\\
			& \equiv -128 \pmod{77}\\
			& \equiv 26 \pmod{77}.
		\end{align*}
		\item
		Let's just verify it modulo $7$ and $11$ which is equivalent
		according to the CRT
		
		\begin{align*}
			c^{43} & \equiv 26^{43} \pmod{7}\\
			& \equiv (-2)^{7 \cdot \varphi(7) + 1} \pmod{7}\\
			& \equiv -2 \pmod{7}\\
			& \equiv 75 \pmod{7}
		\end{align*}
		and
		
		\begin{align*}
			c^{43} & \equiv 26^{43} \pmod{11}\\
			& \equiv 4^{4 \cdot \varphi(11) + 3} \pmod{11}\\
			& \equiv 4^{3} \pmod{11}\\
			& \equiv 64 \pmod{11}\\
			& \equiv -2 \pmod{11}\\
			& \equiv 75 \pmod{11}.
		\end{align*}
	\end{enumerate}
\end{solution}


%TP3 Ex2
\begin{solution}
	\begin{enumerate}
		\item
		\begin{enumerate}
			\item
			$\epsilon_7$ should be a multiple of $11$ so there is a $y$ such
			that $\epsilon_7 = 11y$.
			So we have
			
			\begin{align*}
				11y & \equiv 1 \pmod{7}\\
				y & \equiv (11)^{-1} \pmod{7}\\
				& \equiv 2 \pmod{7}
			\end{align*}
			which gives $\epsilon_7 \equiv 22 \pmod{77}$.
			\item
			$\epsilon_{11}$ should be a multiple of $7$ so there is a $y$ such
			that $\epsilon_{11} = 7y$.
			So we have
			
			\begin{align*}
				7y & \equiv 1 \pmod{11}\\
				y & \equiv 7^{-1} \pmod{11}\\
				& \equiv 8 \pmod{11}
			\end{align*}
			which gives $\epsilon_{11} \equiv 56 \pmod{77}$.
			\item
			By the superposition principle
			\begin{align*}
				5\epsilon_7 + 9\epsilon_{11} & \equiv 5 \pmod{7}\\
				5\epsilon_7 + 9\epsilon_{11} & \equiv 9 \pmod{11}
			\end{align*}
			so $x \equiv 5 \cdot 22 + 9 \cdot 56 \equiv 75 \pmod{77}$.
		\end{enumerate}
		\item
		We can take $x_p\epsilon_7 + x_q\epsilon_{11}$
		since by the superposition principle
		\begin{align*}
			x_p\epsilon_7 + x_q\epsilon_{11} & \equiv x_p \cdot 1 + x_q \cdot 0 \pmod{7}\\
			& \equiv x_p \pmod{7}\\
			x_p\epsilon_7 + x_q\epsilon_{11} & \equiv x_p \cdot 0 + x_q \cdot 1 \pmod{11}\\
			& \equiv x_q \pmod{11}.
		\end{align*}
		\item
		We can know that $\epsilon_p = qy$ for some $y$.
		So we have
		\begin{align*}
			qy & \equiv 1 \pmod{p}\\
			y & \equiv q^{-1} \pmod{p}\\
		\end{align*}
		which gives $\epsilon_7 \equiv qq^{-1} \pmod{pq}$
		where the inverse is taken modulo $p$.
		
		This is the same for $\epsilon_q$.
		\item
		\[ \psi^{-1} : \mathbb{Z}_p \times \mathbb{Z}_q \to \mathbb{Z}_n; (x_p, x_q) \to x_p\epsilon_p + x_q\epsilon_q. \]
	\end{enumerate}
\end{solution}


%TP3 Ex3
\begin{solution}
	\begin{enumerate}
		\item
		It is a subset of $\mathbb{Z}_p^*$ since $0$
		cannot be the square of an element in $\mathbb{Z}_p^*$.
		
		It is a group since if $x_1$ and $x_2$ are quadratic residues,
		there exists $y_1,y_2$ such that
		\begin{align*}
			y_1^2 & \equiv x_1 \pmod{p}\\
			y_2^2 & \equiv x_2 \pmod{p}
		\end{align*}
		so
		\[ (y_1y_2)^2 \equiv x_1x_2 \pmod{p} \]
		and $x_1x_2$ is also a quadratic residue.
		\item
		Let's show that there is one square root between $0$ and $\frac{p-1}{2}$ and one between $\frac{p-1}{2}+1$ and $p-1$.
		Using a primitive root $g$ of $p$, and $\alpha,\beta$ such that $g^\alpha = a$ and $g^\beta = b$,
		\begin{align*}
			a^2 & \equiv b^2 \pmod{p}\\
			g^{2\alpha} & \equiv g^{2\beta} \pmod{p}\\
			2\alpha & \equiv 2\beta \pmod{p-1}\\
			\alpha & \equiv \beta \pmod{\frac{p-1}{2}}.
		\end{align*}
		Hence the uniqueness of the square root between each half.
		A quadratic residue has therefore either 1 or 2 roots.
		But if $a^2 \equiv x \pmod{p}$, we also have $(-a)^2 \equiv x \pmod{p}$ and $a \not\equiv -a \pmod{p}$ since $p$ is odd and $a \neq 0$.
		As a consequence, it cannot have only 1 root.
		\item
		If all quadratic residues has 2 roots.
		Since there is $p$ roots and $a$ cannot be the root of 2 different quadratic residues,
		there must be $\frac{p-1}{2}$ quadratic residues (since we do not consider 0).
		The probability is therefore $1/2$.
		\item
		If $x$ is a QR, there is $y$ such that $y^2 \equiv x \pmod{p}$.
		We have
		\begin{align*}
			x^{\frac{p+1}{2}}
			& \equiv x^{\frac{p-1}{2}}x \pmod{p}\\
			& \equiv y^{p-1}x \pmod{p}\\
			& \equiv x \pmod{p}
		\end{align*}
		The other one is $-x^{\frac{p+1}{4}}$.
		If both of them is a quadratic residue, then $-1 = \frac{-x^{\frac{p+1}{4}}}{x^{\frac{p+1}{4}}}$ is one too.
		We know that one of its square root is $(-1)^{\frac{p+1}{2}} = 1$ since $\frac{p+1}{2}$ is even which is absurd
		since $1^2 \not\equiv -1 \pmod{p}$.
		\item
		It is obvious that if $0 \leq a < p-1$ is even, $g^a$ is a QR.
		Since only half are QR, $g^a$ cannot be a QR if $a$ is odd.
		$g^a$ is therefore a QR iff $a$ is even.
		
		$g^{ab}$ is a QR if $a$ or $b$ are even or equivalently
		if $g^a$ or $g^b$ are QR.
		If $g^a$ and $g^b$ are odd, $h_b$ cannot be a QR.
		If it is we are sure that $b = 0$.
		
		Let build an adversary $\A$ that outputs 1 if $1$ if the parity of
		$ab$ is consistent with the parity of $a$ and $b$ and $0$ otherwise.
		\begin{itemize}
			\item if $b = 0$, we win iff $c$ has a different parity than $ab$, so the probability of winning $1/2$.
			\item if $b = 1$, we win.
		\end{itemize}
		Therefore
		\[ \Pr[\DDH_{\A,\G}(n) = 1] = \frac{1}{2} + \frac{1}{4} = \frac{3}{4}. \]
		DDH is solved since $\frac{1}{4}$ is non-negligible.
	\end{enumerate}
\end{solution}


%TP3 Ex4
\begin{solution}
	\begin{enumerate}
		\item
		We have
		\[ \epsilon_7 + \epsilon_{11} \equiv 78 \equiv 1 \pmod{77} \]
		and
		\[ \varphi(1) = (1,1) = (1,0) + (0,1). \]
		We also have
		\[ \epsilon_7 \cdot \epsilon_{11} \equiv (2 \cdot 11) \cdot (8 \cdot 7) \equiv 16 \cdot 77 \equiv 0 \pmod{77} \]
		and
		\[ \varphi(0) = (0,0) = (1,0) \cdot (0,1). \]
		\item           
		We have: $\epsilon_p = q(q^{-1} \pmod{p}$ and $\epsilon_q = p(p^{-1} \pmod{q}$, so, $\epsilon_p + \epsilon_q = q(q^{-1} \pmod{p}) + p(p^{-1} \pmod{q})$ where p and q are relatively prime.
		
		
		And also $\epsilon_p + \epsilon_q = q u + p v$ where u and v are the inverses of p (resp q). We have $q u + p v = 1 \pmod{pq}$ (see Bézout theorem).
		We also have $\epsilon_p \epsilon_q = pq(q^{-1} \pmod{p})(p^{-1} \pmod{q}) = 0 \pmod{pq}$. Whe then have a generalization:
		
		
		Finally: $\varphi(1) = (1, 1)$ and $\varphi(0) = (0, 0)$.
		
		\item
		For the sum whe have: $x = k_1 p + x_p$ and $y = k_2 p + yp$, then whe have $x + y = p(k_1 + k_2) + (x_p + y_p)$.
		
		
		For the multiplication: $\varphi(x)\varphi(y) = (x_p, x_q)(y_p, y_q) = (x_p y_p, x_qy_q) = ((xy)_p, (xy)_q) = \varphi(xy)$.
	\end{enumerate}
	
	
	The decryption can take advantage of it since it knows $p$ and $q$.
\end{solution}


%TP3 Ex5
\begin{solution}
	\begin{enumerate}
		\item
		If $x \in QR(n)$, there is $y$ such that $y^2 \equiv x \pmod{n}$.
		We have seen that $(y_p^2,y_q^2) = \varphi(y)^2 = \varphi(y^2) = \varphi(x)$
		so $x$ is also a QR modulo $p$ and $q$.
		
		If $x \in QR(p)$ and $x \in QR(q)$, there is $r_1,r_2$ such that
		$r_1^2 \equiv x_p \pmod{p}$ and $r_2^2 \equiv x_q \pmod{q}$.
		We have seen that
		$x = \varphi^{-1}(x_p,x_q) = \varphi^{-1}(r_1^2,r_2^2) = \varphi^{-1}(r_1,r_2)^2$
		so $x \in QR(n)$.
		
		If $F$ is a injection, it is also a bijection since it maps $QR(n)$ to itself.
		Let's prove that it is an injection.
		If $x^2 \equiv y^2 \pmod{p}$, $x_p^2 \equiv y_p^2 \pmod{p}$ so
		$(x_p-y_p)(x_p+y_p) \equiv 0 \pmod{p}$.
		Since $p$ is prime, we have either
		$(x_p-y_p) \equiv 0 \pmod{p}$
		or
		$(x_p+y_p) \equiv 0 \pmod{p}$
		The second one is impossible as we have seen in the exercise 3 since that would mean that $-1 \in QR(p)$
		which is absurd since $p \equiv 3 \pmod{4}$.
		We have therefore $x_p \equiv y_p \pmod{p}$,
		the same reasoning gives $x_q \equiv y_q \pmod{q}$ so $x \equiv y \pmod{n}$.
		\item
		Half the numbers are in $QR(p)$ and half are in $QR(q)$ so $1/4$ are in $QR(n)$.
		So each elements of $QR(n)$ has 4 roots.
		These are simple obtained by using the CRT on $\pm$ the square root modulo $p$
		and $\pm$ the square root modulo $q$.
		For example, with 36, we have $6$ for both modulo 7 and 11.
		The 4 square roots modulo $n$ are therefore $\varphi^{-1}(\pm 1, \pm 5)$.
		\begin{align*}
			\epsilon_7 + 5\epsilon_{11} & \equiv -6 \pmod{77}\\
			\epsilon_7 - 5\epsilon_{11} & \equiv -27 \pmod{77}\\
			-\epsilon_7 - 5\epsilon_{11} & \equiv 6 \pmod{77}\\
			-\epsilon_7 + 5\epsilon_{11} & \equiv 27 \pmod{77}.
		\end{align*}
		$F^{-1}(36)$ is the one that is in $QR(n)$ so in $QR(p)$ and $QR(q)$.
		We see that $1^2 \equiv 1 \pmod{7}$ and $4^2 \equiv 5 \pmod{16}$ so $F^{-1}(36) = -6$.
		\item
		\begin{enumerate}
			\item
			We would have $n | (x_0-x_1)(x_0+x_1)$.
			However, since $x_0 \not\equiv \pm x_1 \pmod{n}$,
			$n \nmid (x_0-x_1)$ and $n \nmid (x_0 + x_1)$.
			Therefore, $\gcd(x_0-x_1,n)$ cannot $n$ and cannot be 1 either
			since that would mean that $n | (x_0 + x_1)$.
			Therefore $\gcd(x_0-x_1,n)$ is $p$ or $q$.
			\item
			Let's pick a random $x_1$ and run the the preimage algorithm on $x_1^2$.
			If it outputs $x_0 = \pm x_1$, restart over.
			The probability of this working at each iteration is $1/2$ since there is exactly $4$ roots,
			we have $1/2$ chance to pick the one that is not outputted by the algo.
		\end{enumerate}
		\item
		For an input message $m$, take
		\[ c := m^2 \pmod{n}. \]
		For the decryption, since there is 4 roots, a mechanism should be provided for the Decryption to know which one it is.
		This is why RSA is more used in practice while it has the desadvantage of having no proof to be as unbreakable as factoring.
		
		It is not CPA secure since it is not randomized.
	\end{enumerate}
\end{solution}


%TP3 Ex6
\begin{solution}
	\begin{enumerate}
		\item
		We know that
		\[ e_{eve}d_{eve} \equiv 1 \pmod(N) \]
		so $e_{eve}d_{eve}-1$ is a multiple of $N$.
		\item
		Let's pick a random $g \in \mathbb{Z}_N^*$ different from $\pm 1$.
		Let $k'$ such that $k = k'\phi(N)$.
		We know that $g^k \equiv 1 \pmod{N}$.
		Considering the sequence
		$g^k, g^{k/2}, g^{k/4}, g^{k/8}, \ldots, g^{k/2^{e_1}}$
		where $e_1$ is the exponent of $2$ in the prime decomposition of $k$.
		We know that
		\begin{align*}
			1
			& = (g^{k/2^{e_1}})^{2^{e_1}} \pmod{N}\\
			((g^{k/2^{e_1}})^{2^{e_1}} - 1) & = 0 \pmod{N}\\
			((g^{k/2^{e_1}})^{2^{e_1-1}} - 1)((g^{k/2^{e_1}})^{2^{e_1-1}} + 1) & = 0 \pmod{N}
		\end{align*}
		If $g^{k/2^{e_1}} \not\equiv \pm 1 \pmod{N}$,
		$N$ divides neither of the 2 factors and we can get $p$ or $q$
		by computing $\gcd((g^{k/2^{e_1}})^{2^{e_1-1}} - 1, N)$.
		
		If $g^{k/2^{e_1}} \equiv \pm 1 \pmod{N}$,
		we need to take another $g$ at random.
		
		Sadly, I have found that this does not necessarily work.
		If $p \equiv q \equiv 3 \pmod{4}$ for example (let's also say $k' = 1$ for simplicity),
		we have
		\begin{align*}
			g^{k} & \equiv 1 \pmod{N}\\
			g^{k/2}
			& \equiv g^{2\frac{p-1}{2}\frac{p-1}{2}} \pmod{N}\\
			& \equiv 1 \pmod{p}\\
			& \equiv 1 \pmod{q}\\
			& \equiv 1 \pmod{N}\\
		\end{align*}
		$g^{k/4}$ cannot be different from $-1$ or $1$ because that would mean that
		we have found another root for $1$ than $\pm 1$ modulo $p$ or $q$.
		\item
		I think I don't understand the question right.
		Charly only sees $N$, $e_{alice}$ and $e_{bob}$
		but $e_{alice}$ and $e_{bob}$ gives him no infomation.
		it is just number relatively prime to $N$,
		we could have computed them just from $N$.
	\end{enumerate}
\end{solution}


% TP4 EX1
\begin{solution}
	\begin{description}
		\item[computationally hiding]
		It has to know whether $c \in \Ima(G)$ and whether $c \xor R \in \Ima(G)$.
		This is easy to do with unbounded power so it is not perfectly hiding.
		
		This is impossible to do in less than $2^{n-1}$ with a random function.
		If $\A(G,R)$ can beat computational hiding, we can build $\A'^g$ that gives $g^{s_1}$ to $\A$.
		If $\A$ outputs 0, $\A'$ says that $g$ is a PRF and if $\A$ outputs 1,
		that means that it is wrong so that may be because $g$ is a random function
		so $\A'$ says that $g$ is a random function.
		\item[computationally biding]
		It is not perfectly biding since we can loop over all $s$ of $\{0,1\}^n$ and find two $s_1,s_2$ such
		that $G(s_1) \xor R = G(s_2)$.
		This is $O(\sqrt{2^n})$ by the Birthday Paradox.
		
		It is however computationally binding.
		Let's suppose that $\A(G,R)$ can output $(c, (s_1,b_1), (s_2, b_2))$ such that
		$G(s_1) \xor \langle b_1,R \rangle = c = G(s_2) \xor \langle b_2,R \rangle$.
		This is impossible with a random function, the best we can do with random function is the technique described just below.
		
		We can therefore build a distinguisher $\D$ which will give the function it is given to $\A$.
		If succeeds, it concludes that it is a PRF, otherwise, it concludes that this is a random function.
	\end{description}
\end{solution}


% TP4 EX2
\begin{solution}
	\begin{enumerate}
		\item
		This scheme is perfectly hiding.
		$r$ is picked at random uniformly in $\mathbb{Z}_N^*$
		so $c(0) = r^{2}$ is uniformly distributed in $QR(N)$ since each element of $QR(N)$ has the same number of roots.
		$c(1) = mr^{2}$ is also uniformly distributed in $QR(N)$
		since this is a subgroup of $\mathbb{Z}_N^*$.
		Therefore both $c(0)$ and $c(1)$ are random elements uniformly distributed in $QR(N)$.
		\item
		Because it's perfectly hiding, it cannot be perfectly binding.
		If there is a PPT adversary $\A$ that can find with non-negligible probability
		$r_{0}$, $r_{1}$ such that $\Open(c, (r_{0}, 0))=0$ and
		$\Open(c, (r_{1}, 1))=1$. This would mean that
		\[ c(0)=c(1) \Leftrightarrow r_{0}^{2}=mr_{1}^{2} \Leftrightarrow m=(r_{0}/r_{1})^{2}. \]
		This means that from $\A$, we can build $\A'$ that find the square roots of $m$ with non-negligible probability.
		As seen from exercise~5 of APE3, this means that we can factor $N$.
		
		Therefore, under the assumption that factoring $N$ is hard, the scheme is computationally binding.
		\item
		If the commiter choose $m$, he has the factorisation of $N$ so it is not binding.
		As we have seen in exercise~3 of APE3, we can find the square root of $m$ modulo $p$ and $q$
		and using the CRT, we obtains its square root modulo $m$.
		It suffices now to take $r_0$ and $r_1$ such that $r_{0}/r_{1}$ is this square root which is easy (take for example $r_1 = 1$).
		\item
		If $m \not\in QR(N)$, it is not perfectly hiding since the other party than the commiter can
		check if $c \in QR(N)$.
		If it is, $b = 0$ and if it is not, $b = 1$.
		
		It is not computationally hiding either since the other party have the factorisation of $N$
		so he can easily check if $c \in QR(N)$ with a PPT algorithm.
		\item
		He can ask, once the commitment has been opened,
		one square root at the other party.
		If he can't give it or if it is not a square root of $m$, he knows that the other party has cheated.
		\item
		\begin{itemize}
			\item $\Gen(1^n)$ sets $pk$ as $(N,e,m)$, where
			\begin{itemize}
				\item $N$ is a RSA modulus generated by $\G_\mathrm{RSA}(1^n)$
				\item $e$ is a element of $\mathbb{Z}_{\phi(N)}^*$
				\item $m$ is a random element of $\mathbb{Z}_{N}^*$
			\end{itemize}
			\item $\Com_{pk}(b)$ with $b \in \{0,1\}$ provides $(c,d)$ where:
			\begin{itemize}
				\item $c = m^br$, $r$ is a random element of $\mathbb{Z}_N^*$
				\item $d := (r, b)$
			\end{itemize}
			\item $\Open_{pk}(c, d)$ outputs $b$ if it can recompute $c$ from $d$ and $pk$, or $\perp$ otherwise
		\end{itemize}
		\begin{enumerate}
			\item
			This scheme is perfectly hiding because $c$ is a random element of $\mathbb{Z}_N^*$.
			The difference here is that $e$ is relatively prime to $\phi(N)$ since $e \in \mathbb{Z}_{\phi(N)}^*$ (this was not the case of 2, since $\phi(N) = (p-1)(q-1)$ is even).
			Therefore, $e$ has an inverse $d$ modulo $\phi(N)$ and every $x \in \mathbb{Z}_N^*$ has a $e$th root which is $x^d$.
			\item
			If there is a PPT adversary $\A$ that can find with non-negligible probability
			$r_{0}$, $r_{1}$ such that $\Open(c, (r_{0}, 0))=0$ and
			$\Open(c, (r_{1}, 1))=1$. This would mean that
			\[ c(0)=c(1) \Leftrightarrow r_{0}^e=mr_{1}^e \Leftrightarrow m=(r_{0}/r_{1})^e. \]
			This means that from $\A$, we can build $\A'$ that find the $e$th roots of $m$ with non-negligible probability.
			This means that $\A'$ can solve RSA problem!
			
			Under the assumption that the RSA problem is hard, the scheme is computationally binding.
			\item
			If the commiter choose $m$, he has the factorisation of $N$ so it is not binding.
			The commiter can find $\phi(N)=(p-1)(q-1)$ and find the inverse $d$ of $e$ modulo $\phi(N)$
			and ``decrypt'' $m$ ``à la RSA''.
			It suffices now to take $r_0$ and $r_1$ such that $r_{0}/r_{1}$ is this ``decryption'' which is easy (take for example $r_1 = 1$).
			\item
			If $m$ is not in $\mathbb{Z}_N^*$, then the commiter can compute $\gcd(m, N)$ and find a factor of $N$, the scheme won't be binding.
			The other party can easily check if $c \in \mathbb{Z}_N^*$ so it won't be hiding either.
			\item
			He can simply check that $\gcd(m, N) = 1$.
		\end{enumerate}
	\end{enumerate}
\end{solution}

% TP4 Ex4
\begin{solution}
	No, they should use a commitment scheme.
	As a commitment point of view, this scheme is perfectly hiding, bit it is not binding.
	For example, if Bob has sent $c_{\text{bob}} := \mathsf{abraveboy} \xor k_{\text{bob}}$ and sees that it is a girl,
	he can say that its key was $c_{\text{bob}} \xor \mathsf{acutegirl}$ and Alice will
	compute $c_{\text{bob}} \xor (c_{\text{bob}} \xor \mathsf{acutegirl}) = \mathsf{acutegirl}$ and think that he guessed right.
\end{solution}

% TP4 Ex5
\begin{solution}
	Let's show that $\A$ can break CDH, we can build $\A'$ that can break DDH.
	\begin{itemize}
		\item $\A'$ is given $\mathbb{G}, q, g, (g^x, g^y, h_b)$.
		\item $\A'$ gives $\mathbb{G}, q, g, g^x, g^y$ to $\A$.
		\item $\A'$ receives $h$ from $\A$.
		\item If $h = h_b$, $\A'$ outputs 1, otherwise, it outputs 0.
	\end{itemize}
	Let's analyse our probability of breaking DDH
	\begin{itemize}
		\item
		If $\A$ succeeds in finding $h = g^{xy}$ the only think that could go wrong
		is if the random $z = xy$ (probability of $1/q$) when $b = 0$ (probability of $1/2$).
		\item
		Even if $\A$ loses and
		\begin{itemize}
			\item $b = 0$, we win if $g^z \neq h$ which happens with probability $1-1/q$;
			\item $b = 1$, we lose.
		\end{itemize}
	\end{itemize}
	In conclusion,
	
	\begin{align*}
		\Pr[\DDH_{\A',\G}(n) = 1]
		& = \Pr[\DDH_{\A',\G}(n) = 1]\left(1 - \frac{1}{2q}\right)\\
		& \quad + (1 - \Pr[\DDH_{\A',\G}(n) = 1])\left(\frac{1}{2}-\frac{1}{2q}\right)\\
		& = \frac{1}{2} + \frac{\Pr[\DDH_{\A',\G}(n) = 1]}{2} - \frac{1}{2q}
	\end{align*}
	where $\frac{\Pr[\DDH_{\A',\G}(n) = 1]}{2} - \frac{1}{2q}$ is non-negligible if $\Pr[\DDH_{\A',\G}(n) = 1]$
	is non-negligible.
\end{solution}


%TP5 Ex4
\begin{solution}
	\begin{enumerate}
		\item
		It is not a secure scheme.
		We can ask for the signature of $m_1$ which is $(\sigma_1, \sigma_2, s)$ and give the existential forgery
		$(m, (\sigma_1, \Mac_s(m), s))$.
		The signature $\sigma_1$ has been replayed
		\item
		It is a secure scheme.
		Let's prove that if $\A$ an produce an existential forgery,
		we can build one for either the regular signature scheme or the one-time signature scheme.
		
		$\A$ does the $q(n)$ queries $m_1, \ldots, m_{q(n)}$ and receive the signatures $(\sigma_{1;i}, \sigma_{2;i}, {vk'}_i)$.
		If it then outputs an existential forgery $(m, (\sigma_1, \sigma_2, vk'))$, let's consider 2 cases
		\begin{itemize}
			\item
			Either $vk' \neq {vk'}_i$ for all $i$.
			Does that mean that $\sigma_2$ is an existential forgery for the one-time signature scheme ?
			We could say ``\emph{Not necessarily since $sk',vk'$ are picked at random, with very bad luck, two same could have been picked.
				And even if they are different, thinking about Lamport, some $y$ could be the same between two keys !}''.
			However here we deal with \emph{any} one-time signing scheme so we cannot reason about Lamport at a lower level and find an invertion of $f$.
			
			Fortunately, this is absolutely not a problem.
			An adversary $\A'$ trying to find an existential forgery for the one-time signature scheme has access $\Sig$,
			it is allowed to pick new random $(sk', vk')$ and compute $\sigma_2$ for those.
			Since it is PPT, it only has the time to pick a polynomial number of those so it is not a problem.
			
			Let's do a reduction where it is clear that there is no problem
			\begin{enumerate}
				\item $\A'$ receives $vk'$ and pick a random $i^* \in \{0, \ldots, q\}$ (he tries to guess for which ${vk'}_i$ will be reused).
				\item $\A'$ gives $pk$ to $\A$
				\begin{itemize}
					\item When $\A$ sends $m_i$ for $i \neq i^*$, $\A'$ pick random $({sk'}_i, {vk'}_i)$ and
					computes sends $(\Sign_{sk}({vk'}_i), \Sig_{{sk'}_i}(m_i), {vk'}_i)$ to $\A$.
					\item When $\A$ sends $m_i$ for $i = i^*$, $\A'$ asks its single query $m_i$ and receives $\sigma_{2;i} = \Sig_{sk'}(m_i)$.
					He computes $\sigma_{1;i} = \Sign_{sk}({vk'}_i)$ and sends $(\sigma_{1;i}, \sigma_{2;i}, vk')$ to $\A$.
				\end{itemize}
				\item $\A$ tries outputs an existential forgery, if it is valid and $vk' = {vk'}_{i^*}$ (probability $1/q(n)$),
				$\A'$ also has an existential forgery and wins.
			\end{enumerate}
			\item
			Either $vk' \neq {vk'}_i$ for all $i$.
			That means that $\sigma_1$ is an existential forgery for the regular signature scheme.
			
			From an adversary $\A$ for this scheme, we
			can build an adversary $\A''$ for the regular signature scheme
			simply outputs $(m,\sigma_1)$ when $\A$ outputs its
			existantial forgery $(\sigma_1, \sigma_2, m)$
			hoping that $vk'$ is not the same as an previous ${vk'}_i$.
			In that case, it will win if $\A$ wins.
		\end{itemize}
		
		In summary, we have
		\begin{align*}
			\Pr[\Sigforge_{\A}(n) = 1]
			& = \Pr[\Sigforge_{\A}(n) = 1 \land (\exists i: vk' = {vk'}_i)]\\
			& \quad + \Pr[\Sigforge_{\A}(n) = 1 \land (\forall i: vk' \neq {vk'}_i)]\\
			& \leq \Pr[\Sigforge_{\A''}(n) = 1] + \frac{\Pr[\Sigforgeone_{\A'}(n) = 1]}{q(n)}
		\end{align*}
		which is non-negligible since $\Pr[\Sigforge_{\A''}(n) = 1]$ and $\Pr[\Sigforgeone_{\A'}(n) = 1]$ are negligible and $q(n)$ is polynomial.
	\end{enumerate}
\end{solution}


%TP7 Ex1
\begin{solution}
	Here, $2^n$ tags are valid for a single message (since $r$ can be anything).
	The problem with this scheme is that the function $\xor: \{0,1\}^n \times \{0,1\}^n \to \{0,1\}^n$
	cannot be injective.
	We have for example, $m_1 \xor m_2 = (m_1 \xor w) \xor (m_2 \xor w)$.
	
	For this scheme, a valid tag of $m_1 \| \cdots \| m_l$ (using $r$)
	is therefore also a valid tag for $(m_1 \xor w) \| \cdots \| (m_l \xor w)$ (using $r \xor w$).
	
	Our adversary $\A$ will simply submit the query $m_1 \| \cdots \| m_l$ and receive the tag $t$.
	It will then submit the forgery $((m_1 \xor w) \| \cdots \| (m_l \xor w), t)$.
\end{solution}


%TP7 Ex3
\begin{solution}
	Let's suppose that $N_1 \neq N_2 \neq N_3$ (it should be in the exercise statement).
	Let's first check that $\gcd(N_1,N_2) = 1$, $\gcd(N_2,N_3) = 1$ and $\gcd(N_3,N_1) = 1$.
	If it is not the case, we can immediately factor one of the $N_i$ and find $\phi(N_i)$, $d_i$ and decrypt $c_i$.
	
	Let's now consider that $N_1, N_2$ and $N_3$ are relatively prime.
	
	We have
	\begin{align*}
		m^3 & \equiv c_1 \pmod{N_1}\\
		m^3 & \equiv c_2 \pmod{N_2}\\
		m^3 & \equiv c_3 \pmod{N_3}
	\end{align*}
	and using the CRT we can find $c$ such that $m^3 \equiv c \pmod{N_1N_2N_3}$.
	
	However, if we assume that $m \leq N_i$ for $i = 1, 2, 3$ to simplify as suggested,
	that means that $m^3 < N_1N_2N_3$ (equality would mean that $N_1=N_2=N_3$ which is not true).
	We have therefore simply the regular equation (no modulo)
	\[ m^3 = c \]
	which can be computed easily (we can solve it simply with binary search for example).
\end{solution}

%TP7 Ex4
\begin{solution}
	We have
	\begin{align*}
		\Enc(m_1m_2)
		& = (m_1m_2)^E\\
		& = m_1^Em_2^E\\
		& = \Enc(m_1)\Enc(m_2).
	\end{align*}
	
	This version of RSA is not randomized so breaking its CPA or CCA secure is completely obvious since we can ask
	the decryption of the same message several times.
	
	However, this is not the case for decryption so if this version was random, we would have to find a smarter attack.
	Using $\Enc(m_1 m_2) = \Enc(m_1)\Enc(m_2)$, we know that $\Dec(\Enc(m_1 m_2)) = \Dec(\Enc(m_1)\Enc(m_2))$ so
	$m_1 m_2 = \Dec(\Enc(m_1)\Enc(m_2))$.
	Hence we can do the query $(m_0,m_1)$ (with $m_0 \neq m_1$), receive $c_b$ and then ask for an encryption of a random $c_2$ of $m_2$ hoping that $c_2 \neq 1$ (if it is not, we pick another $m_2$ and start over).
	Afterwards, we can ask for the decryption of $c_bc_2$ which is allowed if $c_2$ is not 1.
	We will get $m_bm_2$, it is now trivial to find $m_b$ and then $b$.
\end{solution}